EDIpack: A generic and interoperable Lanczos-based solver for quantum impurity problems

### 2025-05-23 03_edipack2.tex
Finally implemented the cg_norm part I updated accordingly the EDIpack code. We might want to further change the code, and so the description here, but I think for now this reads as honest and clear.

In the superconducting case, where {\tt ed\_mode}={\bf superc}, 
special care is needed to properly handle the anomalous (off-diagonal) 
components within the Nambu basis. In this case, the bath functions 
must include both the normal and anomalous components, 
which can significantly alter the form of the self-energy.

\subsection{Bath Optimization}\label{sSecFit}
In the DMFT framework, the bath parameters must be optimized to 
faithfully reproduce the Weiss field, 
${\GG^{-1}_{0}}_{\a\b\s\s'}(z)$, or its corresponding hybridization 
function 
\begin{equation}
\Theta(z) = (z + \mu)\11 - H^\mathrm{loc} - \GG_0^{-1}(z),
\end{equation}
where $\mu$ is the chemical potential and $H^\mathrm{loc}$ is the 
local non-interacting Hamiltonian. The Weiss field is obtained from the DMFT 
self-consistency equation~\cite{Georges1996RMP}, while the bath 
discretization requires a careful fitting of this continuous function 
using a finite set of bath parameters (see \secu{sSecBath}).

Several algorithms have been proposed for this optimization 
step~\cite{Garcia2004PRL,Taranto2012PRB}, each with strengths suited 
to different physical contexts. To maintain flexibility in \NAME, 
we provide a general optimization interface that is independent 
from the core impurity solver. However, we also include a fully 
integrated optimization routine based on the conjugate gradient (CG) 
minimization of the cost function:
\begin{equation}
\chi = \sum_{n=1}^{L_\mathrm{fit}} 
\frac{1}{W_n} \left\|X(i\omega_n) - 
X^\mathrm{QIM}(i\omega_n; \{V, h\}) \right\|_q,
\end{equation}
where the $q$-norm is a suitably chosen distance metric in the 
matrix function space. Here, $X_{\a\b} = {\GG_{0}}_{\a\b\s\s'}$ or 
$\Theta_{\a\b\s\s'}$ are the user-supplied local functions, while 
$X_{\a\b}^\mathrm{QIM} = G^0_{\a\b\s\s'}$ or 
$\Delta_{\a\b\s\s'}$ are the corresponding functions for the quantum 
impurity model, as defined in \secu{sSecFunc}.

@@

To provide maximum control over the fitting process, we include several 
tunable parameters:
\begin{itemize}
\item \texttt{cg\_method=0,1} - Chooses the CG algorithm: 
\texttt{0} for a Fletcher-Reeves-Polak-Ribiere variant adapted from 
Numerical Recipes~\cite{NumRec77}, and \texttt{1} for the original 
algorithm described in Ref.\onlinecite{Georges1996RMP}, commonly used 
in the DMFT community.
\item \texttt{cg\_grad=0,1} - Sets the gradient calculation method: 
\texttt{0} for analytical gradients (when available), \texttt{1} for 
numerical gradients (required if \texttt{cg\_method=1}).
\item \texttt{cg\_Lfit} - Sets the number of Matsubara frequencies 
$L_\mathrm{fit}$ used in the fit. This can be used to restrict the 
fit to the low-frequency regime, where the function behavior is 
most relevant.
\item \texttt{cg\_Weight=0,1,2} - Determines the frequency weighting 
scheme: \texttt{0} for uniform, \texttt{1} for inverse frequency 
weighting, and \texttt{2} for inverse Matsubara index weighting, which 
can emphasize low-energy contributions.
\item \texttt{cg\_pow} - Sets the power $q$ of the cost function, 
allowing the user to fine-tune the sensitivity of the optimization 
to outliers.
\item \texttt{cg\_Ftol} - Controls the fit tolerance, setting the 
convergence threshold for the CG minimization.
\item \texttt{cg\_Niter} - Specifies the maximum number of allowed 
iterations for the CG minimization.
\item \texttt{cg\_stop=0,1,2} - Defines the exit condition of the 
minimization. The options are:
\begin{align*}
C_1 & = |\chi^{n-1} - \chi^n| < \mathtt{cg\_Ftol} (1+\chi^n), \\
C_2 & = \left\|x_{n-1} - x_n\right\| < 
\mathtt{cg\_Ftol} (1+\left\|x_n\right\|),
\end{align*}
where $\chi^n$ is the cost function at the $n^\mathrm{th}$ step, and 
$x_n$ the corresponding parameter vector. The stop condition is a 
logical OR between these criteria, with the parameter value selecting 
which condition to apply.
\end{itemize}

@@

% In the DMFT context the bath needs to be optimized
% against a given realization of the Weiss field
% ${\GG^{-1}_{0}}_{\a\b\s\s'}(z)$, or the corresponding {\it hybridization} function
% $\Theta = (z + \mu)\11 - H^{loc} -\GG_0^{-1}](z)$. While the Weiss
% field is obtained from the DMFT self-consistency
% equation~\cite{Georges1996RMP}, bath discretization impose to  find an optimal
% description of the continuous Weiss field in terms of the finite
% number of parameters of the bath (see \secu{sSecBath}.
% Although different algorithms can be envisaged to perform this
% step~\cite{Garcia2004PRL,Taranto2012PRB} which are perfectly valid in
% some physical contexts.
% In order to enable the users to use their preferred 
% method in \NAME we keep the optimization implementation independent
% from the rest of the code. Yet, we offer a fully fledged optimization
% strategy based on the  conjugate gradient (CG) minimization of the cost function:
% $$
% \chi = \sum_{n=1}^{L_\mathrm{fit}}\frac{1}{W_n}||X(i\omega_n) - X^\mathrm{QIM}(i\omega_n;\{V,h\})||_q
% $$
% where the symbol $|| \cdot ||_q$ is a suitably defined $q$-norm in the
% matrix function space, $X_{\a\b}={\GG_{0}}_{\a\b\s\s'},\,\Theta_{\a\b\s\s'}$ are the user
% supplied local functions and
% $X_{\a\b}^\mathrm{QIM}=G^0_{\a\b\s\s'},\,\Delta_{\a\b\s\s'}$ the
% quantum impurity model functions, defined above.
% Notably, more sophisticated optimization algorithms have been
% recently implemented to better undertake the minimization of similar
% cost functions. However, these developments are designed to work with truncated
% algorithms, e.g. sCI or DMRG, describing systems with a larger number of
% bath levels.   
% In the ED context however, where the efficient and wise selection of
% the relatively few bath levels becomes crucial, the CG minimization of
% the local functions proved to be the most efficient and
% flexible optimization scheme so far.

% The fit procedure is entirely wrapped by the single function {\tt
%   ed\_chi2\_fitgf} contained in the module \texttt{ED\_BATH\_FIT}.   
% In order to exploit regularity of the functions the fit is 
% performed using imaginary Matsubara frequency. 
% The actual form of $X_{\a\b\s\s'}$ is controlled by the input parameter
% \texttt{cg\_Scheme=Weiss,Delta}.

% In order to fine tune the optimization step in \NAME\ we
% introduced a number of control parameters. To start with, we include
% two distinct CG algorithms in the library, controlled by the input variable
% \texttt{cg\_method=0,1}. The value \texttt{0} corresponds to a
% Fletcher-Reeves-Polak-Ribiere minimisation algorithm, adapted from
% Numerical Recipes~\cite{NumRec77}. Yet, in order to guarantee
% back-compatibility with respect to the past literature, we also provide the
% possibility of using the minimization procedure published in
% Ref.\onlinecite{Georges1996RMP} which has been largely used in the DMFT
% community.

% The gradient $\nabla\chi$ required by the CG algorithm can be
% evaluated either analytically or numerically as
% controlled by the value of the input parameter \texttt{cg\_grad=0,1}. For
% \texttt{cg\_method=1} only numerical gradient can be used.
% The number of Matsubara frequency $L_\mathrm{fit}$ used in the fit procedure
% is controlled  by the input parameter \texttt{cg\_Lfit}.
% This parameter can be used to restrict the fit to the low frequency
% part as the large frequency tails of the local Matsubara functions
% have universal behavior.
% Similarly, the weight  $W_n=1,1/L_\mathrm{fit},1/\omega_n$ is used to
% enhance the weight of low energy part with respect to the
% intermediate-to-large energy one in the cost function.
% This is controlled by the input parameter \texttt{cg\_Weight=0,1,2}.
% Another  factor contributing to determine the behavior
% of the fit is the power $q$ of the cost function $\chi$. By tuning
% this parameter, controlled by input variable \texttt{cg\_pow}, it is
% possible to enhance or suppress the differences in the $\chi$ in order
% to improve (or simplify) the optimization procedure.

% Finally, we introduced few additional parameters which aim to
% control the quality of the fit. The first is the fit tolerance,
% controlled by the input parameter \texttt{cg\_Ftol}, which
% determines the convergence threshold of the minimization procedure.
% The second is the maximum number of iterations \texttt{cg\_Niter}
% allowed for the CG algorithm to determine the minimum.
% The last parameter \texttt{cg\_stop=0,1,2} selects the exit
% condition of the minimization procedure. We envisaged  three
% possible conditions: $C_1\cup C_2$, $C_1$, $C_2$ with 
% $C_1=|\chi^{n-1} -\chi^n|<\mathtt{cg\_Ftol}(1+\chi^n)$ and
% $C_2=||x_{n-1} -x_n||<\mathtt{cg\_Ftol}(1+||x_n||)$, where $\chi^n$
% is the cost function evaluated at the $n^{\rm th}$-step and $x_n$
% the corresponding argument. Increasing the value of the stop
% parameter generally loosens up the exit conditions of the minimization.

{\bf REMARK TODO}
Notes on the norm function: a) the definition for conventional
matrices, ;b) the definition for the replica/general case where you
can choose between different norm defition.
\fixme{After careful inspection, I (GB) realize that the elemental norm 
was never really ported from CDFMT-ED to EDIpack (there's no different
matrix weight than all ones, which is the standard chi2, of course). 
Hence the \texttt{cg\_norm} options is really not doing anything significant,
as with "elemental" it gives the usual $\chi^2$ (or $\chi^q$, if you change the
\texttt{cg\_pow}...) and for "Frobenius" it gives again the usual $\chi^2$,
unless you put $q\neq2$, which gives an ill-defined matrix norm...
(it mixes an effective $q=2$ coming from the matrix multiplication with 
$q\neq2$ when you do the explicit $q$-root). I think we
should just get rid of \texttt{cg\_norm} and do the traditional $\chi^q$, as 
indeed we are already doing by default. Implemented like this, it is just random
no-consequence code.}
%GBtodo placeholder

>>>>

In the superconducting case, where {\tt ed\_mode}={\bf superc}, 
special care is needed to properly handle the anomalous (off-diagonal) 
components within the Nambu basis. In this case, the bath functions 
must include both the normal and anomalous components, 
which can significantly alter the form of the self-energy. Similar issues arise in the {\bf nonsu2} case, where spin off-diagonal components usually drive the physics of the problem.

\subsection{Bath Optimization}\label{sSecFit}
In the DMFT-ED framework, the bath parameters must be optimized to 
faithfully reproduce the Weiss field, 
${\GG^{-1}_{0}}_{\a\b\s\s'}(z)$, or its corresponding hybridization 
function 
\begin{equation}
\Theta(z) = (z + \mu)\11 - H^\mathrm{loc} - \GG_0^{-1}(z),
\end{equation}
where $\mu$ is the chemical potential and $H^\mathrm{loc}$ is the 
local non-interacting Hamiltonian. The Weiss field is obtained from the DMFT 
self-consistency equation~\cite{Georges1996RMP}, while the bath 
discretization requires a careful fitting of this continuous function 
using a finite set of bath parameters (see \secu{sSecBath}).

Several algorithms have been proposed for this optimization 
step~\cite{Garcia2004PRL,Taranto2012PRB,Mejuto_Bath_ASCI-DMFT}, each with strengths suited 
to different physical contexts. To maintain flexibility in \NAME, 
we provide a general optimization interface that is independent 
from the core impurity solver. However, we also include a fully 
integrated optimization routine based on the conjugate gradient (CG) 
minimization of the cost function:
\begin{equation}
\chi = \sum_{n=1}^{L_\mathrm{fit}} 
\frac{1}{W_n} \left\|X(i\omega_n) - 
X^\mathrm{QIM}(i\omega_n; \{V, h\}) \right\|_q,
\label{eq:chiq}
\end{equation}
where the $q$-norm is a suitably chosen distance metric in the 
matrix function space. Here, $X_{\a\b} = {\GG_{0}}_{\a\b\s\s'}$ or 
$\Theta_{\a\b\s\s'}$ are the user-supplied functions of the bath parameters, while 
$X_{\a\b}^\mathrm{QIM} = G^0_{\a\b\s\s'}$ or 
$\Delta_{\a\b\s\s'}$ are the corresponding quantities for the quantum 
impurity model, as defined in \secu{sSecFunc}.

@@

To provide maximum control over the fitting process, we include several 
tunable parameters:
\begin{itemize}
\item \texttt{cg\_method=0,1} - Chooses the CG algorithm: 
\texttt{0} for a Fletcher-Reeves-Polak-Ribiere variant adapted from 
Numerical Recipes~\cite{NumRec77}, and \texttt{1} for the original 
algorithm described in Ref. \cite{Georges1996RMP}, commonly used 
in the DMFT community.
\item \texttt{cg\_grad=0,1} - Sets the gradient calculation method: 
\texttt{0} for analytical gradients (when available), \texttt{1} for 
numerical gradients (required if \texttt{cg\_method=1}).
\item \texttt{cg\_Lfit} - Sets the number of Matsubara frequencies 
$L_\mathrm{fit}$ used in the fit. This can be used to restrict the 
fit to the low-frequency regime, where the function behavior is 
most relevant.
\item \texttt{cg\_Weight=0,1,2} - Determines the frequency weighting 
scheme: \texttt{0} for uniform, \texttt{1} for inverse frequency 
weighting, and \texttt{2} for inverse Matsubara index weighting, which 
can emphasize low-energy contributions.
\item \texttt{cg\_pow} - Sets the power $q$ of the cost function, 
allowing the user to fine-tune the sensitivity of the optimization 
to outliers.
\item \texttt{cg\_norm=elemental,frobenius} - Sets the matrix norm to be used to define the cost function in \equ{eq:chiq}. The default value (\texttt{elemental}) corresponds to a generic element-wise $\chi^q$ norm, defined as 
\begin{equation*}
    ||X-X^\mathrm{QIM}||_q \equiv {\sum_{ij} \left|X_{ij}-X^\mathrm{QIM}_{ij}\right|^q}.
\end{equation*}
The \texttt{frobenius} option implements instead the matrix distance induced by the Frobenius inner product as
\begin{equation*}
    ||X-X^\mathrm{QIM}||_q \equiv 
    \sqrt[q]{\Tr\left[(X-X^\mathrm{QIM})^\dagger(X-X^\mathrm{QIM})\right]} =
    \left(\sum_{ij} \left|X_{ij}-X^\mathrm{QIM}_{ij}\right|^2 \right)^{\!\!\!\frac{1}{q}}.
\end{equation*}
For $q=2$ (see \texttt{cg\_pow}) the Frobenius norm is equivalent
to the Euclidean distance in matrix space. 
%For $q\neq2$ we are implementing something that is not a matrix norm (better to not comment on it explicitly)
We prospect to expand the choices for \texttt{cg\_norm} in future updates of \NAME, as the \textbf{nonsu2} and \textbf{superc}
diagonalization modes entail nontrivial subtleties in optimizing 
the off-diagonal components of $X$.
\item \texttt{cg\_Ftol} - Controls the fit tolerance, setting the 
convergence threshold for the CG minimization.
\item \texttt{cg\_Niter} - Specifies the maximum number of allowed 
iterations for the CG minimization.
\item \texttt{cg\_stop=0,1,2} - Defines the exit condition of the 
minimization. The options are:
\begin{align*}
C_1 & = |\chi^{n-1} - \chi^n| < \mathtt{cg\_Ftol} (1+\chi^n), \\
C_2 & = \left\|x_{n-1} - x_n\right\| < 
\mathtt{cg\_Ftol} (1+\left\|x_n\right\|),
\end{align*}
where $\chi^n$ is the cost function at the $n^\mathrm{th}$ step, and 
$x_n$ the corresponding parameter vector. The stop condition is a 
logical OR between these criteria, with the parameter value selecting 
which condition to apply.
\end{itemize}

@@

### 2025-05-23 05_examples.tex
Finally implemented the cg_norm part I updated accordingly the EDIpack code. We might want to further change the code, and so the description here, but I think for now this reads as honest and clear.

\section{Examples}\label{SecExamples}
In this section we illustrate the functioning of the \NAME
library and their interfaces as solver for DMFT through specific examples, including
discussion of relevant code snippets.

>>>>

\section{Examples}\label{SecExamples}
In this section we illustrate the functioning of the \NAME
library and their interfaces as solver for DMFT through a variety of examples, including
discussion of relevant code snippets.
